---
title: '241 Project (Summer 2017): Effect of Meditation on Blood Pressure'
author: "Thong Bui, Erika Lawrence, Matt Post"
output:
  pdf_document: default
  html_document: default
---
## Source code
Our entire project code and documents can be accessed from https://github.com/thongnbui/MIDS_241_project. This file is can be found at code/ProjectOhm.{Rmd,pdf}. Data file is data/meditation.csv

##Abstract:
We examine the effects of meditation on blood pressure and pulse rates. In particular, using a within-subject experimental design, we compare the effects of a single fifteen-minute meditation session to a similar session of coloring graphic shapes. Subjects were randomly assigned to undergo either the meditation session first, followed by coloring, or the coloring session first, followed by meditation. In each case, a five-minute washout period was inserted between the sessions.  

Generally, we find that both meditation and coloring have some lowering effect on blood pressure and pulse. Details of our analysis will be further explored below.

##Introduction:
Meditation has been used for thousands of years by religious practitioners for spiritual growth. In modern times, it is promoted as a means to reduce stress. Many scientific studies have shown its effectiveness in reducing anxiety and other negative states. 

###What is our question? 
Our study seeks to take this a step further: is it meditation that decreases stress, or is it activites that accompany meditation that are effective? Can other calming activities work as well as meditation to reduce stress? We chose coloring as the activity that most closely resembles traditional meditation without the meditation part. Coloring is a quiet, focused activity, that is not verbal. We chose geometric shapes as a backdrop to avoid the focus on people or cultural associations. Similarly, in meditation there is a distancing of the connection to thoughts and people, and a focus on the abstract. By comparing similar activities, we can narrow our search to those activities that best reduce stress.

###Why does it matter?
High blood pressure, left untreated, can harm a person's health for years before symptoms develop. Patients with high blood pressure may become disabled or suffer a fatal heart attack. About half of the people with high blood pressure who are left untreated die of heart disease, and another third die of stroke.  

####Systolic vs Diastolic Blood Pressure vs Pulse
Our study uses portable wrist blood pressure monitors to measure three numbers: systolic and diastolic blood pressure, and pulse rates. Of these numbers the most important are systolic blood pressure (the first number given when reporting blood pressure), and pulse.  Diastolic pressure is generally only used during an emergency such as a cardiac arrest. Pulse is more variable than systolic pressure, but both are indicators of stress levels. 

##Experiment

###Overview
Our experiment observed thirty-two people both in groups and individually, in person and online. Subjects varied in age from under ten years to over eighty, and in years of meditation experience from zero to fifty. These volunteers were primarily drawn from researcher friends and family networks. 

Our subjects were randomly divided into two groups: one group that listened to a fifteen-minute guided meditation (from https://www.appropriateresponse.com/teachings/) first and spent fifteen minutes coloring after, and another group that engaged in the coloring exercise first, followed by the guided meditation. Between each session, subjects were asked to get up and walk around for 5 minutes, to create a 'washout' period and avoid spillover effects. 

Subjects were asked to measure their own blood pressure before and after each session, and the results were recorded either by the subject (during group sessions) or by the researcher (for individual sessions) after each measurement. 

###Considerations

#####Why within-subject design
The within-subject design gave the study two advantages: first, using the same person to represent both experimental and control conditions gives greater precision around treatment estimates, and second, we were able to use fewer volunteers for the same statistical power.

Within-subject designs have a few potential sources of bias: treatment assignment, anticipation, and persistence. We used an R function to assign subjects to either meditation-first or coloring-first groups. As new volunteers signed up, they were added to the next slot on a list which automatically assigned the pre-randomized group. 

To avoid anticipation, subjects were not told what the sessions would entail until the sessions began. There may have been some guesswork, due to our pre-experiment questions that identified years of meditation experience, or to the packages that arrived in the mail that included colored pencils. However, the details were not known to the subjects in advance. 

A washout of five minutes was used to combat persistence. While it might have been better to take a longer period of time to 'reset' our subjects blood pressures, we tried to keep this time short in consideration of the overall time our volunteers were giving to the study.  

#####Why Coloring
Coloring is used quite often in medical experiments as a control activity.  We used it because of its similarity to meditation - subjects are focused, seated, not interacting with other people, not required to engage in verbal processing, and silent. It was the closest activity we found that included many of the calming features of meditation, without actually being meditation. 

#####Why a Guided Meditation
We chose an online guided meditation from www.appropriateresponse.com for the meditation sessions. There was a great deal of debate about this choice within the team: meditation styles include silent, guided, mantras, visualizations, music and more. In the end, it was decided that for beginners, a guided meditation would be less arduous. Mantra meditations might be distracting, especially in group settings. In a music meditation, separating out whether it was the music or the meditation that created the effects would be impossible. And visualization meditations might take too long to explain, and have their own set of confounding factors (who gives the instructions, if they would be remembered, etc.).  

Guided meditations themselves come with an array of variation. The meditation we chose was deemed to have the least potentially negative reaction, as it avoids spiritual and esoteric language. However, some of the more experienced meditators who were accustomed to a different mediation style reported that they found the guided meditation distracting.

###Challenges

#####BP measurement device precision
While we were pleased to find affordable mobile blood pressure monitors, wrist monitors are not the best measuring devices for precision. Each person required some basic instructions in the use of the monitors - in most cases this was the first time they used a device of this kind, which may have increased the likelihood of measurement error. In one case we were forced to remove the measured results, as they were deemed an error by a medical professional.

#####Subject inexperience
Experienced meditators often aim for 30 to 60 minutes in a given session. However, this length of time might be uncomfortable for inexperienced meditators, so we limited the sessions to 15 minutes.  An interesting follow-up study might test the results of a longer meditation, using only subjects with meditation experience. 

#####Subject participation
The original study design called for subjects to test the long-term results of meditation. For practicality and do avoid non-compliance or attrition, the design was altered to measure the results of a single meditation. A follow up study might take advantage of subjects on a meditation retreat to look for longer term effects, although the generalizability of such findings would certainly be called into question.

##Data:
```{r libraries, message=FALSE, warning=FALSE}
library(data.table)
library(sandwich)
library(lmtest)
library(stargazer)
library(RColorBrewer)
library(cobalt)
library(car)
```

###Data Exploration

#####Read in and clean up the data
```{r data_clean}
d <- fread("../data/meditation.csv")
# Should be 32 subjects
cat("Number of rows:", nrow(d), "\n")
#str(d)
d <- data.table(d)

# ******Clean up the data*****
# 1. We are remove ID = 13 because this person's BP number is likely machine error
# 2. Reverse group number: 1 should be meditation first aka "med-first", 
#   0 is coloring 1st aka "color-first"
# 3. convert string categories to numeric

d2 <- d[ID != 13 , ID, .(Group = ifelse(Group == 0, 1, 0), #Exclude 13th person & reverse groups
            Recruited_By,   caffeinated_drinks, Age_Group,     Gender,
            Religion,       Years_practice,     hours_since_last_caffeinated_drink,
            previous_strenuous_activity,        Before_Meditation_how_relaxed,
            Post_Med_focus, Enjoy_Coloring,
            in_person = ifelse(Online_in_person == 'I', 1, 0),
            pre_existing_BP = Pre_existing_blood_pressure,
            b4_all_sys = ifelse(Group == 0, B4_Med_BP_Sys, B4_color_BP_Sys),
            b4_all_pulse = ifelse(Group == 0, B4_Med_BP_PUL, B4_color_BP_Pul),
            B4_Med_BP_Sys, B4_Med_BP_PUL,
            After_Med_BP_Sys, After_Med_BP_PUL,
            B4_color_BP_Sys, B4_color_BP_Pul, 
            After_color_BP_Sys, After_color_BP_Pul) ]

#summary(d2)
```

#####Chart the data

```{r histograms, fig.height=3}
par(mfrow=c(1,2))
hist(d2$Years_practice, breaks = 30,col="blue", 
     main = "Histograms: Years of Practice", xlab = "Years")
hist(d2$hours_since_last_caffeinated_drink, breaks = 30, col="green",
     main = "Hours since Caffeine", xlab = "Hours")
hist(d2$B4_Med_BP_Sys, breaks = 30, xlim=c(80,140), col="orange",
     main = "Histograms: Prior Syst.", xlab = "Systolic Pressure")
hist(d2$After_Med_BP_Sys, breaks = 30, xlim=c(80,140),col="gold",
     main = "After Meditation Syst.", xlab = "Systolic Pressure")
hist(d2$B4_color_BP_Sys, breaks = 30, xlim=c(80,140), col="purple",
     main = "Histograms: Prior Pulse", xlab = "Pulse")
hist(d2$After_color_BP_Sys, breaks = 30, xlim=c(80,140), col="violet",
     main = "After Meditation Pulse", xlab = "Pulse")
```

```{r expl_charts, fig.height=3.7}
cols<-brewer.pal(n=3,name="Set1")
cols_t1<-cols[d2$Group+1] # red for group 0, blue for 1
med_diff <-  d2$B4_Med_BP_Sys - d2$After_Med_BP_Sys
color_diff <-   d2$B4_color_BP_Sys - d2$After_color_BP_Sys
par(mfrow =c(1,2))
plot(d2$Group ~ med_diff , col=cols_t1, pch=16, ylab="Group")
legend("center",legend=c("Group 1: med-first", "Group 0: color-first"), 
       col=cols_t1, pch=16)
plot(d2$Group ~ color_diff , col=cols_t1, pch=16, ylab="Group")
legend("center",legend=c("Group 1: med-first", "Group 0: color-first"), 
       col=cols_t1, pch=16)

# Plot b4 vs after meditation for 2 groups
plot(After_Med_BP_Sys ~ B4_Med_BP_Sys, data=d2 , col=cols_t1, pch=16, 
     xlab = "Before Meditation", ylab="After Meditation", main="Before-After Correlation M")
legend(x=98,y=200, legend=c("Group 1: med-first", "Group 0: color-first"), 
       col=cols_t1, pch=16, xpd=TRUE)

# Plot b4 vs after coloring for 2 groups
plot(After_color_BP_Sys ~ B4_color_BP_Sys, data=d2 , col=cols_t1, pch=16, 
     xlab = "Before Coloring",ylab="After Coloring", main = "Before-After Correlation C")
legend(x=100,y=175, legend=c("Group 1: med-first", "Group 0: color-first"), 
        col=cols_t1, pch=16, xpd=TRUE)

```


###Covariate Balance Check
Matching was performed using the Matching package (Sekhon, 2011), and covariate balance was assessed using cobalt (Greifer, 2017), both in R 3.3.0 (R Core Team, 2016).

When using bal.tab() with continuous treatments, the balance statistic presented is the (weighted) Pearson correlation between each covariate and treatment.

```{r cov_balance, warning=FALSE}
covs <- subset(d2, select = 
                 -c(Group, After_Med_BP_Sys, After_Med_BP_PUL, 
                    After_color_BP_Sys, After_color_BP_Pul))

d2$p.score <- glm(f.build("Group", covs), data = d2, 
                       family = "binomial")$fitted.values
d2$att.weights <- with(d2, Group + (1-Group)*p.score/(1-p.score))

bal.tab(covs, treat = d2$Group, weights = d2$att.weights,
         method = "weighting", estimand="ATE")
```
#####Covariate Balance Concerns
Of the covariates analyzed, the following stand out as unbalanced: caffeinated drinks, religion, and years of practice. The Med-First group is more likely to have participants who drink caffeine, are Buddhist and have more meditation experience, while the Color-First group is more likely to include subjects that do not identify with a particular religion. 

The difference in prior blood pressure and pulse measures is perhaps even more concerning. Systolic blood pressure measures before coloring for color-first participants tended to be consistently lower than for those starting out with meditation. 

In each case, it will be important to include these variables when running regressions, to account for any possible bias.

```{r balance_plots, fig.width=4,fig.height=3}
bal.plot(covs, treat = d2$Group, weights = d2$att.weights, method = "weighting",
         estimand = "ATT", var.name = "caffeinated_drinks")
bal.plot(covs, treat = d2$Group, weights = d2$att.weights, method = "weighting",
         estimand = "ATT", var.name = "Years_practice")
bal.plot(covs, treat = d2$Group, weights = d2$att.weights, method = "weighting",
         estimand = "ATT", var.name = "Religion")
bal.plot(covs, treat = d2$Group, weights = d2$att.weights, method = "weighting",
         estimand = "ATT", var.name = "B4_Med_BP_PUL")
bal.plot(covs, treat = d2$Group, weights = d2$att.weights, method = "weighting",
         estimand = "ATT", var.name = "B4_color_BP_Sys")
```

#####Feature creation and selection

```{r feature_create, warning=FALSE}
#create bins for meditation experience
d2 <- transform(d2, experience=cut(Years_practice,  
              breaks=c(-1,1, 10, 25, 50), labels= c(0,10, 25, 50)))
d2 <- transform(d2, exp_bin=cut(Years_practice,  
              breaks=c(-1,0, 50),  labels= c(0,1)))
d2 <- transform(d2, pre_med_sys=cut(B4_Med_BP_Sys, 
              breaks = c(50, 100, 125, 150), labels = c("low", "med", "high")))
d2 <- transform(d2, pre_col_sys=cut(B4_color_BP_Sys, 
              breaks = c(50, 100, 125, 150), labels = c("low", "med", "high")))
d2 <- transform(d2, pre_med_pul=cut(B4_Med_BP_PUL, 
              breaks = c(50, 70, 90, 110),  labels = c("low", "med", "high")))
d2 <- transform(d2, pre_col_pul=cut(B4_color_BP_Pul, 
              breaks = c(50, 70, 90, 110), labels = c("low", "med", "high")))

age_str = "c('5-9', '10-19') = '< 20'; '30-39' = '30s';"
age_str = paste(age_str , "'40-49'= '40s'; c('50-59', '60-69', '80-90') = '50+'")
d2$age = recode(d2$Age_Group, age_str)

d2$prev_act = recode(d2$previous_strenuous_activity, "'No' = 'no'; else='yes'")

#create fields to capture the difference before and after each session
#for clarity, we subtract the number we expect to be smaller (after) 
#from the larger (before), to keep the numbers positive
#Sysolic
d2$med_sys_diff = d2$B4_Med_BP_Sys - d2$After_Med_BP_Sys
cat("Average Meditation Effect on Systolic Blood Pressure:", mean(d2$med_sys_diff), "\n")
d2$col_sys_diff = d2$B4_color_BP_Sys - d2$After_color_BP_Sys
cat("Average Coloring Effect on Systolic Blood Pressure:", round(mean(d2$col_sys_diff),2), "\n")
#Pulse
d2$med_pul_diff = d2$B4_Med_BP_PUL - d2$After_Med_BP_PUL
cat("Average Meditation Effect on Pulse Rate:",round(mean(d2$med_pul_diff),2), "\n")
d2$col_pul_diff = d2$B4_color_BP_Pul - d2$After_color_BP_Pul
cat("Average Coloring Effect on Pulse Rate:",round(mean(d2$col_pul_diff),2), "\n")
# The Coloring ATE is higher than Meditation, so subtract meditation results from coloring
# Create difference in differences measures between meditation and coloring
cat("\nAverage Difference on Systolic Blood Pressure between Coloring and Meditation Effects:", 
    round(mean(d2$col_sys_diff - d2$med_sys_diff), 2), "\n")

cat("\nAverage Difference on Pulse Rate between Coloring and Meditation Effects:", 
    round(mean(d2$col_pul_diff - d2$med_pul_diff), 2))

```

#####Correlations and Complications

```{r corr_compl}
cat("Are Recruiter, Age and Experience correlated?\n\n")
cat("Recruiter by Subject Age Group\n")
table(d2$Recruited_By, d2$age)
cat("\nExperience by Subject Age Group\n")
table(d2$experience, d2$age)

cat("\nATE between meditation vs color by Group. Did we leave enough time for a washout?\n\n")

ate_group_1_diff = mean(d2$col_sys_diff[d$Group==TRUE], na.rm=TRUE) - 
   mean(d2$med_sys_diff[d2$Group==TRUE], na.rm=TRUE)
cat("   Med-first Group Systolic Difference:", round(ate_group_1_diff, 2), "\n")

ate_group_0_diff = mean(d2$col_sys_diff[d$Group==FALSE], na.rm=TRUE) - 
   mean(d2$med_sys_diff[d2$Group==FALSE], na.rm=TRUE)
cat("   Color-first Group Systolic Difference:", round(ate_group_0_diff, 2), "\n")


ate_group_1_diff = mean(d2$col_pul_diff[d$Group==TRUE], na.rm=TRUE) - 
   mean(d2$med_pul_diff[d2$Group==TRUE], na.rm=TRUE)
cat("   Med-first Group Pulse Difference:", round(ate_group_1_diff, 2), "\n")

ate_group_0_diff = mean(d2$col_pul_diff[d$Group==FALSE], na.rm=TRUE) - 
   mean(d2$med_pul_diff[d2$Group==FALSE], na.rm=TRUE)
cat("   Color-first Group Pulse Difference:", round(ate_group_0_diff, 2), "\n")


```
Based on these results, it would be best to include either Age, or Experience or Recruiter in a model because they are correlated.  Always include Group, as the washout period may have been less effected than we might have hoped.

###Model Building:

#####Meditation Effects:


```{r med_mod}
#Systolic Blood Pressure
med_sys_exp_ols = lm(med_sys_diff ~ Group + pre_med_sys + experience + Gender + 
               Religion +  caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus , data=d2)

med_sys_age_ols = lm(med_sys_diff ~ Group + pre_med_sys + age + Gender + 
               Religion + caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus , data=d2)

#Pulse
med_pul_exp_ols = lm(med_pul_diff ~ Group + pre_med_pul + experience + Gender + 
               Religion + caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus , data=d2)

med_pul_age_ols = lm(med_pul_diff ~ Group + pre_med_pul + age + Gender + 
               Religion + caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus , data=d2)

```

#####Color Effects:
```{r col_mod}
col_sys_exp_ols = lm(col_sys_diff ~ Group + pre_col_sys + experience +
               Religion + caffeinated_drinks + in_person + 
               prev_act  + Enjoy_Coloring , data=d2)

col_sys_age_ols = lm(col_sys_diff ~ Group + pre_col_sys + age + 
               Religion + caffeinated_drinks + in_person + 
               prev_act  + Enjoy_Coloring , data=d2)

#Pulse
col_pul_exp_ols = lm(col_pul_diff ~ Group + pre_col_pul + experience +
               Religion + caffeinated_drinks + in_person + 
               prev_act + Enjoy_Coloring  , data=d2)

col_pul_age_ols = lm(col_pul_diff ~ Group + pre_col_pul + age +
               Religion + caffeinated_drinks + in_person + 
               prev_act   + Enjoy_Coloring , data=d2)

```

#####Compare the Meditation and Color effects linear models, separately
```{r stargazer_ind}
#compare meditation models
cat("Meditation Model Comparison\n")
stargazer(med_sys_exp_ols, med_sys_age_ols, med_pul_exp_ols, med_pul_age_ols, type = "text")
#compare coloring models
cat("\n\nColor Model Comparison\n")
stargazer(col_sys_exp_ols, col_sys_age_ols, col_pul_exp_ols, col_pul_age_ols, type = "text")
```

### Simpler models
1. From the original data table, each row:

+-----------+-----------+----------+--------------+-------------+----------------+
| person_id | b4_all_bp |b4_med_bp | after_med_bp | b4_color_bp | after_color_bp |
+===========+===========+==========+==============+=============+================+
|1          | 121       | 121      |   111        |   120       |110             |
+-----------+-----------+----------+--------------+-------------+----------------+

is convert into 5 rows like this: 

+-----------+-----+--------+-----------+-----------+----------+-------------+
| person_id | bp  | b4_all | b4_med    | after_med | b4_color | after_color |
+===========+=====+========+===========+===========+==========+=============+
|1          |121  |   1    |   0       |0          |0         | 0           |
+-----------+-----+--------+-----------+-----------+----------+-------------+
|1          |121  |   0    |   1       |0          |0         | 0           |
+-----------+-----+--------+-----------+-----------+----------+-------------+
|1          |111  |   0    |   0       |1          |0         | 0           |
+-----------+-----+--------+-----------+-----------+----------+-------------+
|1          |120  |   0    |   0       |0          |1         | 0           |
+-----------+-----+--------+-----------+-----------+----------+-------------+
|1          |110  |   0    |   0       |0          |0         | 1           |
+-----------+-----+--------+-----------+-----------+----------+-------------+

then build the model `bp ~ 1 + b4_all + after_med + after_color + person_id`

```{r}
d2.bp = melt(d2, id.vars = c("ID"), measure.vars = c("b4_all_sys", "B4_Med_BP_Sys", "After_Med_BP_Sys","B4_color_BP_Sys", "After_color_BP_Sys"), variable.name = "bp_type", value.name = "bp")
#d2.bp$b4_all <- ifelse(d2.bp$bp_type == "b4_all_sys", 1, 0)
d2.bp$b4_med <- ifelse(d2.bp$bp_type == "B4_Med_BP_Sys", 1, 0)
d2.bp$after_med <- ifelse(d2.bp$bp_type == "After_Med_BP_Sys", 1, 0)
d2.bp$b4_color <- ifelse(d2.bp$bp_type == "B4_color_BP_Sys", 1, 0)
d2.bp$after_color <- ifelse(d2.bp$bp_type == "After_color_BP_Sys", 1, 0)
#summary(d2.bp)
```

Now, we build the same data table for pulse data
```{r}
d2.pulse = melt(d2, id.vars = c("ID"), measure.vars = c("b4_all_pulse", "B4_Med_BP_PUL", "After_Med_BP_PUL","B4_color_BP_Pul", "After_color_BP_Pul"), variable.name = "pulse_type", value.name = "pulse")
#d2.pulse$b4_all <- ifelse(d2.pulse$pulse_type == "b4_all_pulse", 1, 0)
d2.pulse$b4_med <- ifelse(d2.pulse$pulse_type == "B4_Med_BP_PUL", 1, 0)
d2.pulse$after_med <- ifelse(d2.pulse$pulse_type == "After_Med_BP_PUL", 1, 0)
d2.pulse$b4_color <- ifelse(d2.pulse$pulse_type == "B4_color_BP_Pul", 1, 0)
d2.pulse$after_color <- ifelse(d2.pulse$pulse_type == "After_color_BP_Pul", 1, 0)
#summary(d2.pulse)
```

2. Now we are building the simple models
```{r}
m1 <- lm(bp ~ 1 + b4_med + b4_color + after_color + after_med + ID, data=d2.bp)
m2 <- lm(pulse ~ 1 + b4_med + b4_color + after_color + after_med + ID, data=d2.pulse)

#coeftest(m1, vcovHC(m1))
stargazer(m1, m2,type="text")
```

##### Now we are seeing the opposite of what we found earlier: 
* both mediation and coloring are both statistically significant at lowering blood pressure!
* both mediation and coloring has no statistical significance at lowering pulse
* What can we conclude from ID?

These models are comparing post-meditation and post-coloring BPs to BP before any exercises start? Is this a valid way to do it? That's why we have different result from the previous models?

##### Why there is such difference? Which ones are more reliable: simpler ones or previous ones?

### Check the significance of each treatment
```{r}
t.test(d2$B4_Med_BP_Sys, d2$After_Med_BP_Sys)
t.test(d2$B4_Med_BP_PUL, d2$After_Med_BP_PUL)
t.test(d2$B4_color_BP_Sys, d2$After_color_BP_Sys)
t.test(d2$B4_color_BP_Pul, d2$After_color_BP_Pul)
```

###Difference-in-differences Estimate
```{r diff}
# Create some Diff variables
d2$col_med_sys_diff = d2$col_sys_diff - d2$med_sys_diff
d2$col_med_pul_diff = d2$col_pul_diff - d2$med_pul_diff
d2$pre_sys_diff= d2$B4_color_BP_Sys - d2$B4_Med_BP_Sys
d2$pre_pul_diff= d2$B4_color_BP_Pul - d2$B4_Med_BP_PUL

cat("Difference in differences on systolic blood pressure:", mean(d2$col_med_sys_diff))
cat("\nDifference in differences on pulse rate:", mean(d2$col_med_pul_diff), "\n")
```

###Linear Regression model with difference-in-difference Estimate
```{r diff_models}

cat("\nDifference in Differences Model Comparison:\n")
dd_sys_ols1 = lm(col_med_sys_diff~ Group + pre_sys_diff + experience + Gender + 
               Religion +  caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus, data = d2)
dd_sys_ols2 = lm(col_med_sys_diff~ Group + pre_sys_diff + age + Gender + 
               Religion +  caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus, data = d2)

dd_pul_ols1 = lm(col_med_pul_diff~ Group + pre_pul_diff + experience + Gender + 
               Religion +  caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus, data = d2)
dd_pul_ols2 = lm(col_med_pul_diff~ Group + pre_pul_diff + age + Gender + 
               Religion +  caffeinated_drinks + in_person + 
               prev_act + Before_Meditation_how_relaxed + Post_Med_focus, data = d2)

stargazer(dd_sys_ols1, dd_sys_ols2, dd_pul_ols1, dd_pul_ols2,type = "text")
```

### Check the significance of the difference-in-differences
```{r}
t.test(d2$col_sys_diff, d2$med_sys_diff)
t.test(d2$col_pul_diff, d2$med_pul_diff)
```

### Results:
Both the coloring treatment and the meditation treatment on average lowered the blood pressure and pulse of our subjects.  Our results show the effect from coloring to be more powerful than meditation. However, t-tests show that this difference is not statistically significant. A number of variables influence the variation in our results:

##### Meditation:
A number of variables were found to influence the results of meditation on systolic blood pressure and pulse.  In both cases, the **Group** was significant, telling us that the washout period was not sufficient to avoid spillover effects. One of the most significant variables influencing systolic blood pressure differences was the **initial systolic blood pressure**: those subjects who began the session with a higher systolic measure were most likely to have a larger difference. **Initial pulse** was significant to a lesser degree on the post-meditation pulse measurement. **Experience** also showed a significant impact on systolic blood pressure in our model, particularly for participants who had 1 to 10 years of meditation experience.  However this is a group of only two people. In addition, those with 10 to 25 years of experience were most likely to see a benefit on meditation to their pulse rates. **Previous activity** appears to be highly significant in influencing systolic blood pressure from meditation, and on pulse to a lesser degree. **Religion** had little effect on pulse, but those who identified with a particular faith showed a lessening effect for meditation on blood pressure. **Caffeinated beverages** were significant for blood pressure only.  Finally, pulse dropped more when the experiment was conducted **in person**, but this variable had a more limited impact on blood pressure. 

One of the surprising results to the researchers was that those who espoused a particular religion, particularly those that are associated with meditation (Buddhism and Hinduism), saw the smallest effect of the meditation session on blood pressure. It may be that, for people who are more accustomed to meditation, the particular format of the guided meditation was distracting, or that a longer meditation would have revealed different results. It is also possible that over time, meditation has less of an effect on a person's blood pressure. Though the experiment is not structured to capture causal effects from these variables, it could be a very interesting subject for a future study or at least help inform how we structure future tests.

#####Coloring:
In contrast, those who identified with a particular **Religion** saw significant positive impact on coloring effects on pulse rates, and to a lesser degree on systolic blood pressure. In this case, those who identified as having no **religion** also saw a positive benefit, while those identifying as agnostic or 'other' did not have a significant effect. As expected, **prior measurements** of blood pressure and pulse were highly significant. **Age** was significant as well; people over 40 were also much more likely to find coloring highly beneficial. In contrast to meditation, **in person** sessions were less beneficial for colorers.  We could again speculate what causes may be at work, and then generate new hypotheses or implement more careful controls in future experiments.  For example, you could speculate that social anxiety took away from the overall calming effects of coloring, when the coloring took place in groups.

##Conclusion

While there are some statistically significant coefficients from the models above to measure the effect of both meditation and coloring on blood pressure, the t-tests indicate that there are no significant differences between the effects of meditation and coloring on blood pressure and pulse rate. This implies that meditation as well as coloring, and potentially some other methods of relaxation may have similar impacts on lowering blood pressure
which is an interesting discovery that therefore we can't make any causal claim on meditation effect on blood pressure.

### Issues/Concerns

* One of the most difficult experiment decisions is what particular form of meditation to choose from and how long the meditation should be. We decided to go with 15-minute guided meditation from https://www.appropriateresponse.com/teachings/ so our analysis and conclusion can only be based on this particular setup.  If we were to move from a 'class experiment' to a 'full experiment', we would definitely need to consider more carefully various forms and lengths of meditation, as well as how we may need to implement further blocking and clustering, across much larger treatment and control populations.
* The subjects consist of our friends and relatives. This sample is not randomized from the general population therefore doesn't represent the population's distribution hence no generalization can be drawn from our experiment even if we find causal effect from it.
* The wash-out period of 5 minutes may not be sufficient to remove the treatment of the first exercise on the second one.
* The on-line person with ID=13 seems to have unreliable blood pressure measurements which indicates a potential precision issue with the blood pressure device. It can also be an issue with how the experiment was conducted on-line.
* Some tests were conducted 'one-on-one', and some were conducted in groups which can lead to bias.
* It was our first experiment so there maybe mistakes or inconsistencies along the way that we are not aware of.


Based on these facts, we believe more future work needs to be done to refine our experiment design to provide more complete controls and more advanced methods across a larger population, in order to more clearly identify any causal effects of meditation on blood pressure.

###Future Research

Our current project suffers from time constraints in setting up and conducting the experiment, financial limitations in acquiring the appropriate measurement devices, and difficulty in finding enough people to participate.  Future research would need to address and overcome all of these challenges.  

Five of the top improvements we would want to make are:

* Random selection from general population with larger size.
* Better blood pressure device and measurement.
* Longer periods and various forms of meditation.
* To address the insufficient washout period, either increase the time allocated to washout, or utilize 2 separate groups or time periods.
* Possibly implement more advanced blocking on covariates like experience and age to account for variance between groups.
